Swinging, Fast and Slow:\\Interpreting variation in baseball swing tracking metrics

### 2025-06-26 reports/arxiv/paper.tex
add in description of random effects plots and insert leftover residual variance

\usepackage{amsfonts, amsmath, authblk, graphicx, natbib}

@@

    Baseball has a long history as fertile ground for statistical analysis of rich datasets. For Major League Baseball (MLB), detailed records are publicly available at the play-by-play level going back to 1912 \citep{retrosheet}. Beginning in 2007, MLB made optical pitch tracking data freely available online
    \citep{fast_what_2010}, enabling unprecedented insights into player performance \citep{swartz_quality_2017}. In 2016, MLB made waves again when they updated their ball tracking technology to incorporate Doppler radar and released batted ball tracking data to the public \citep{arthur_new_2016}. Most recently, in May 2024, MLB released their most exciting new dataset since batted ball tracking 2016: bat tracking at the pitch-by-pitch level \citep{petriello_everything_2024}.

    \begin{figure}
      \centering
      \includegraphics[width = 0.8\textwidth]{../../figures/counterintuitive.pdf}
      \caption{\it Distribution of bat speed relative to batter average by contact quality (squared up or not), across all swings in the dataset. The x-axis represents the difference between the bat speed and the batter's average bat speed. Following MLB's definition, a swing is considered ``squared up'' if the batted ball's exit velocity is at least 80\% of the theoretical maximum (given pitch speed and bat speed), a proxy for good contact.}
      \label{fig:counterintuitive}
    \end{figure}

    \begin{figure}
      \centering
      \includegraphics[width = 0.4\textwidth]{../../figures/swing_late.pdf}
      \includegraphics[width = 0.4\textwidth]{../../figures/swing_early.pdf}
      \caption{\it A diagram illustrating the effect of swing timing on swing length measurement due to the fact that swing length is measured at the point of contact. Imagining two swings with the exact same mechanics, the left image shows the swing length measurement if the swing is late, and the right image shows the swing length measurement if the swing is early.}
      \label{fig:swing-diagram}
    \end{figure}

@@

      \begin{itemize}
        \item Bat tracking
        \begin{itemize}
          \item \citet{nevins_sensitivity_2019}
          \item \citet{orishimo_lower_2024}
          \item \citet{nakashima_acceptable_2025}
        \end{itemize}
        \item Bat tracking (blogs)
        % Candidates
        % https://blogs.fangraphs.com/what-statcasts-new-bat-tracking-data-does-and-doesnt-tell-us/
        % https://blogs.fangraphs.com/daddy-hacks-or-the-lone-peril-of-swinging-too-hard/
        % https://www.baseballprospectus.com/news/article/95688/best-of-bp-in-defense-of-isaac-paredes/
        % https://www.drivelinebaseball.com/2024/07/using-mlb-bat-tracking-data-to-better-understand-swings/
        % https://theadvancescout.substack.com/p/whats-in-a-swing-a-metrics-explainer
        % https://theadvancescout.substack.com/p/the-radar-gun-doesnt-work-here
        % https://theadvancescout.substack.com/p/the-cheat-swing
        \begin{itemize}
            \item 
        \end{itemize}
        \item Skew normal
        \begin{itemize}
          \item \citet{judge_exit_2024}
        \end{itemize}
        \item Instrumental variable regression
        \begin{itemize}
          \item \citet{putman_tackling_2025}
        \end{itemize}
      \end{itemize}

@@

    Our dataset comprises 685,143 pitches thrown during the 2024 MLB regular season. This publicly available dataset was downloaded from Baseball Savant \citep{baseball_savant} using the R package sabRmetrics \citep{sabRmetrics}. For each pitch, we observe the identities and handedness (left/right) of the batter and pitcher involved and the ball-strike count prior to the pitch. We also observe pitch tracking information, most notably the pitch type (e.g. fastball, curveball, changeup) and the $(x, z)$ location where the pitch crosses the front of home plate. The pitch tracking information is sufficient to recreate an estimate of the pitch's full trajectory as a three-dimensional quadratic function of time $(x(t), y(t), z(t))$ and to recreate the corresponding pitch metrics such as release point, release speed, and vertical/horizontal break. Each pitch has a categorical outcome, which can be one of: hit by pitch, called ball, called strike, swinging strike, foul ball, or hit into play.

@@

@@

    We begin with the hypothesis that two covariates explain real (non-artifactual) differences in swing mechanics: ball-strike count and pitch location. To the extent that a batter’s swing tracking metrics co-vary with count, we describe this as their \textit{approach}. To the extent that a batter’s swing tracking metrics co-vary with pitch location, we describe this as swing \textit{adaptation}. Acknowledging that swing timing biases the measurement of bat speed and swing length, we mitigate this confounding bias by filtering on swings that result in squared-up contact against pitchers' primary fastballs. Additionally, based on visualizations of batter swing length and bat speed distributions (such as examples in INSERT FIGURE REFERENCE?) we are also interested in capturing variation in the shape of the intended distributions for hitters.

@@

    We fit both bat speed and swing length models in a Bayesian framework via the brms package in R \citep{brms}, which provides an interface for Bayesian modeling with Stan \citep{carpenter2017stan}. We use weakly informative priors for the parameters, with vague half-$t_3$ priors (i.e., a Student's $t$ distribution centered at zero with 3 degrees of freedom but truncated to positive values) for the standard deviation parameters \citep{gelman2006prior} and the LKJ prior for the correlations between batter-level effects \cite{lewandowski2009generating}. Our Bayesian approach naturally provides uncertainty quantification for the model parameters via their posterior distributions, which are estimated using MCMC sampling. For model fitting, we use four parallel chains with 6,000 (3,000 burn-in) and 4,000 (2,000 burn-in) samples for the swing length and bat speed models respectively. We observe evidence of convergence of the MCMC algorithm based on observing trace plots and $\hat{R}$ statistics close to 1 \citep{gelman1992inference, brooks1998general}, along with no indication of problematic effective sample sizes \citep{gelman2020bayesian}.

@@

      The intention model from Section \ref{sec:methods-intention} sets us up to estimate a causal model for the effect of bat speed and swing length on swing outcomes via instrumental variables regression (CITATION). Relying on baseball domain knowledge, we hypothesize that the relationship between swing tracking metrics and outcomes is confounded by a latent, unobserved variable: whether the batter correctly identifies the pitch type early in its trajectory (by lucky guess or by other means) or needs to make a mid-swing adjustment. When the batter fails to correctly identify the pitch type early, he is less likely to make good contact, and he also adjusts his swing, reducing the bat speed and affecting the swing length. Critically, this confounder can vary significantly from pitch to pitch as batters frequently limit their focus to one or two pitch types because it is too difficult to prepare for any pitch type (CITATION).

      A good instrumental variable is one which has an effect on the treatment (in this case, swing tracking metrics) and is known to be uncorrelated with the confounder (CITATION). As the instrument varies, we observe the effect of the treatment with the effect of the confounder stripped away. Relying on the framework from Section \ref{sec:methods-intention}, we choose the batter's count-intended bat speed and swing length as our instrumental variables. Count-intended swing length is the player specific contribution of count to swing length, as estimated by (\ref{eqn:intention-swing-length}), and count-intended swing length is defined analogously.

      Count-intended swing metrics affect observed swing metrics as modeled in Section \ref{sec:methods-intention}, and we expect that they are approximately uncorrelated with early pitch recognition. The count is known before the pitch and does not depend on the pitch itself. This may not be a perfect instruction, for example, if batters who tend to struggle more with early pitch recognition in two-strike counts tend also to modify their two-strike swing intentions a certain way. But this choice of instrumental variables will mitigate any confounding effect of early pitch recognition from pitch to pitch.

      We consider the effect of bat speed and swing length on three components of the pitch outcome model: probability of contact conditioned on swing; probability of fair ball conditioned on contact; and expected xwOBA conditioned on fair ball. For each of these components, we estimate a logistic or linear regression of the outcome on the batter's count-intended bat speed and swing length, with an offset for the prediction from the pitch outcome model described in Section \ref{sec:methods-pitch-outcome-model}. This offset controls for the difficulty of each pitch when estimating the effects of count-based bat speed and swing length adjustments.

      Using $i \in \{1, ..., n\}$ to index swings, the random variables $Y_i^{\mbox{\scriptsize con}} \in \{0, 1\}$, $Y_i^{\mbox{\scriptsize fair}} \in \{0, 1\}$ and $Y_i^{\mbox{\scriptsize xwOBA}} \in \mathbb{R}$ represent whether swing $i$ results in contact; whether swing $i$ results in a fair ball; and the xwOBA of the batter ball on swing $i$ (undefined if $Y_i^{\mbox{\scriptsize fair}} \in \{0, 1\} = 0$), respectively. The corresponding predictions from the pitch outcome model are denoted by $\hat p_i^{\mbox{\scriptsize con}} \in (0, 1)$,  $\hat p_i^{\mbox{\scriptsize fair}} \in (0, 1)$ and $\hat y_i^{\mbox{\scriptsize xwOBA}} \in \mathbb{R}$. We use $b_i$ to denote the batter on swing $i$, and $\hat\gamma_{b_i}^{\mbox{\scriptsize BS}}$ and $\hat\gamma_{b_i}^{\mbox{\scriptsize SL}}$ are the estimated batter random slopes for strikes in the bat speed intention model and the swing length intention model, respectively.

@@

@@

      The models from Section \ref{sec:methods-causal} estimate the effect of swing adjustments on contact and power, but they don't dhelp us make value judgments when these effects move in opposite directions, e.g. sacrificing power for increased contact. In baseball, the units of value are runs, which translate directly to wins (CITATION). The standard metric for estimating the run value of batting performance is linear weight, which assigns a run value to each type of batter outcome (strikeout, walk, home run, etc.) according to its average change in base-out run expectancy (CITATION). In this section, we describe a method for converting the effects on contact and power into effects on linear weight.

@@

\begin{table}
        \centering
        \begin{tabular}{l|r|r|r|}
          & $\Delta$ELPD  & SE & \# SEs \\
          \hline
          \input{../../tables/intention_elpd.tex}
        \end{tabular}
        \caption{\it Model comparison via expected log predictive density (ELPD). Shown are the difference in EPLD between the skew-normal (SK) and Gaussian models ($\Delta$ELPD), the standard error of ELPD difference (SE), and the number of standards errors the SK model is better than the Gaussian model.}
        \label{tab:intention-model-elpd}
      \end{table}

@@

      \begin{figure}
        \centering
        \includegraphics[width = 0.6\textwidth]{../../figures/approach.pdf}
        \caption{\it Estimated batter random slopes for strikes in the intention model. Each point is a batter $b$, with the x-value representing $\hat\beta^S + \hat\gamma^S_b$ from the model (\ref{eqn:intention-swing-length}) and the y-value representing the same quantity from the model (\ref{eqn:intention-bat-speed}). This quantity is interpretable as the expected change in swing length (or bat speed) corresponding to a one-strike increase in the ball-strike count.}
        \label{fig:approach}
      \end{figure}

      \begin{table}
        \centering
        \begin{tabular}{l|rrr|rrr|}
                  & \multicolumn{3}{c|}{Bat Speed}          & \multicolumn{3}{c|}{Swing Length}  \\
        Parameter & Mean  & Lower  & Upper & Mean  & Lower  & Upper \\
          \hline
          \input{../../tables/fixed_effects.tex}
        \end{tabular}
        \caption{\it Posterior mean and 95\% credible interval lower/upper bounds for the fixed effects in the bat speed and swing length intention models.}
        \label{tab:fixed-effects}
      \end{table}

>>>>

\usepackage{amsfonts, amsmath, authblk, float, graphicx, natbib, tikz}
\usetikzlibrary{decorations.pathmorphing}

@@

    Baseball has a long history as fertile ground for statistical analysis of rich datasets. For Major League Baseball (MLB), detailed records are publicly available at the play-by-play level going back to 1912 \citep{retrosheet}. Beginning in 2007, MLB made optical pitch tracking data freely available online
    \citep{fast_what_2010}, enabling unprecedented insights into player performance \citep{swartz_quality_2017}. In 2016, MLB made waves again when they updated their ball tracking technology to incorporate Doppler radar and released batted ball tracking data to the public \citep{arthur_new_2016}. In May 2024, MLB released their most exciting new dataset since batted ball tracking in 2016 by making bat tracking data available at the pitch-by-pitch level \citep{petriello_everything_2024}.

    The new bat tracking metrics include {\it bat speed}, {\it fast-swing rate}, {\it squared-up rate}, {\it blasts} and {\it swing length}. Of these, only bat speed and swing length represent new information: The other metrics (fast-swing rate, squared-up rate and blasts) are functions of bat speed and other ball tracking data that were already publicly available. Both bat speed and swing length are measured at the moment of bat-ball contact (or at the point nearest to contact, in the case of a swing-and-miss). Bat speed is the linear speed of the ``sweet spot'' of the bat, approximately six inches from the end; and swing length is the distance travelled by the end of the bat from the start of the swing until contact. Critically, bat speed and swing length are available for each swing, not just aggregated at the player level. This opens the door to pitch-level analysis within player, not just analysis of differences between player averages.

    An open research question in baseball is the effect of batter aggressiveness. Conventional coaching wisdom would suggest that the harder a player swings, the less likely they are to make solid contact, but the harder they are likely to hit the ball if they do make solid contact. What is the effect of bat speed on contact? Figure \ref{fig:counterintuitive} illustrates a simple analysis, and the result is counterintuitive. We observe that swings which result in solid contact actually have slightly faster-than-average bat speeds (relative to the batter's own average). This shows that bat speed is positively correlated with making solid contact. Taken at face value, it would seem to suggest that batters should always swing as hard as possible because it does not reduce contact quality and may even increase it.

    \begin{figure}[H]
      \centering
      \includegraphics[width = 0.8\textwidth]{../../figures/counterintuitive.pdf}
      \caption{\it Distribution of bat speed relative to batter average by contact quality (squared up or not), across all swings in the dataset. The x-axis represents the difference between the bat speed and the batter's average bat speed. Following MLB's definition, a swing is considered ``squared up'' if the batted ball's exit velocity is at least 80\% of the theoretical maximum (given pitch speed and bat speed), a proxy for solid contact.}
      \label{fig:counterintuitive}
    \end{figure}

    Like the famous problem of the bat and ball whose total cost is \$1.10 \citep{kahneman_thinking_2011}, this is a problem for which slower, more deliberative thinking is helpful. The dilemma is due to how the data are measured. The measurement occurs at the point of contact, for which the timing is closely related to the outcome of the swing. Figure \ref{fig:swing-diagram} illustrates two hypothetical swings with identical swing mechanics, but because one swing is early and the other is late, the measurements are taken at different points in the swing. Thus, the timing of the swing affects the measured bat speed and swing length. From this perspective, bat speed and swing length are themselves outcomes of the swing, not entirely under the batter's control.

    \begin{figure}[H]
      \centering
      \includegraphics[width = 0.4\textwidth]{../../figures/swing_late.pdf}
      \includegraphics[width = 0.4\textwidth]{../../figures/swing_early.pdf}
      \caption{\it A diagram illustrating the effect of swing timing on swing length measurement due to the fact that swing length is measured at the point of contact. Imagining two swings with the exact same mechanics, the left image shows the swing length measurement if the swing is late, and the right image shows the swing length measurement if the swing is early.}
      \label{fig:swing-diagram}
    \end{figure}

    We hypothesize that the counterintuitive results shown in Figure \ref{fig:counterintuitive} are explained by an unobserved confounder: the extent to which the batter correctly identifies the pitch type early in its trajectory (by lucky guess or by other means). For example, when the batter prepares for a high fastball and this guess is correct, they are likely to complete their swing with full intent. When the batter preparing for a high fastball is met with a curveball, they need to make a mid-swing adjustment, reducing their bat speed in the process. The latter scenario is less likely to result in contact than the former.

    Our approach to this problem is to model the {\it intent} of the batter rather than taking observed bat speed and swing length at face value. We control for contact quality and we model the distribution of swing metrics for each batter {\it when they are successful} to better identify what they are trying to do. We estimate separate Bayesian hierarchical skew-normal models for intended bat speed and intended swing length, and we demonstrate heterogeneity between batters in how they modulate their intended swing metrics as the number of strikes increases. We perform instrumental variables regression to estimate the causal effect of these intended swing metrics on swing outcomes. Finally, we model a plate appearance as a Markov chain to evaluate the tradeoff between contact and power when modulating bat speed and swing length.

@@

      Shortly after MLB released the new bat tracking data, there were several analyses published in online outlets. \citet{clemens_what_2024} explored correlations on the player level between the new bat tracking metrics and existing performance metrics, finding that the new metrics did not explain much variation in overall performance. This was followed by more careful analyses at the swing level. \citet{woodward_radar_2024} estimated a generalized additive model (GAM) to predict bat speed (as a percentage of the batter's peak bat speed) based on pitch location and other variables; and then manually created rules based on the residuals of this model to label certain individual swings as ``guess'' swings or ``catch-and-redirect'' swings in an effort to identify batter intent. \cite{orr_defense_2024} pointed out that swings resulting in pulled fly balls tend to have greater swing lengths because the contact point tends to be further in front of home plate; and then fit a GAM to predict swing length from pitch location and bat speed in an effort to isolate contact depth.

      In the academic literature, we have not found any previous studies analyzing newly available bat tracking data or examining the relationship between swing aggressiveness and batting outcome in baseball. \citet{nevins_sensitivity_2019} modeled the relationship between bat speed and batted ball speed for fastpitch softball batters. \citet{orishimo_lower_2024} investigated correlations between kinetic/kinematic features and bat speed, finding that ground force reactions under the lead foot (vertical, anterior/posterior, and resultant) were most correlated with bat speed. In perhaps the most intriguing and closely related study to the present work, \citet{nakashima_acceptable_2025} collected full-path bat tracking data from 145 swings made by 29 collegiate batters in a laboratory setting and calculated for each swing the acceptable range of timing error for contacting a pitch with the ``sweet spot'' of the bat. In a different sport, \citet{burton_linear_2015} developed a model for the relationship between volleyball serving agressiveness and service outcomes and calculated optimal aggressiveness based on a server's rates of aces and errors.

      Recent research on statistics in sports has used some of the methodological pieces in the present work. \citet{judge_exit_2024} presented a skew-normal likelihood for the distribution of a batter's exit velocity in baseball, which motivated our approach in Section \ref{sec:methods-intention}. \citet{putman_tackling_2025} used instrumental variable regression (as we do in Section \ref{sec:methods-causal}) to model the causal effect of pass/run play calling on offensive play outcomes in American football.

@@

    Our dataset comprises 685,143 pitches thrown during the 2024 MLB regular season. This publicly available dataset was downloaded from Baseball Savant \citep{baseball_savant} using the \texttt{R} package \texttt{sabRmetrics} \citep{sabRmetrics}. For each pitch, we observe the identities and handedness (left/right) of the batter and pitcher involved and the ball-strike count prior to the pitch. We also observe pitch tracking information, most notably the pitch type (e.g. fastball, curveball, changeup) and the $(x, z)$ location where the pitch crosses the front of home plate. The pitch tracking information is sufficient to recreate an estimate of the pitch's full trajectory as a three-dimensional quadratic function of time $(x(t), y(t), z(t))$ and to recreate the corresponding pitch metrics such as release point, release speed, and vertical/horizontal break. Each pitch has a categorical outcome, which can be one of: hit by pitch, called ball, called strike, swinging strike, foul ball, or hit into play.

@@

    The initial dataset includes attempted bunts (which are labelled) and attempted {\it check-swings} (i.e., partial swings, not labelled). We aim to limit our analysis to full swings. When reporting aggregate metrics like average bat speed, MLB filters out each batter's bottom 10\% of swings by bat speed, labelling the remaining swings as {\it competitive}. We take a more inclusive approach by filtering out only swings with bat speed below 50 mph (which filters out all bunt attempts). Based on video review, this cutoff does a good job of distinguishing between full and partial swings while only removing $\approx$ 2.5\% of swings.

    We reproduce MLB's derived metric for {\it squared-up} contact as:
    \begin{equation}
        \label{eqn:squared-up}
    \mbox{batted ball speed} > 80\% \times \left\{1.23 \cdot (\mbox{bat speed}) + 0.23 \cdot (\mbox{pitch speed})\right\}.
    \end{equation}
    The right-hand side of Equation \ref{eqn:squared-up} is a rough approximation for 80\% of the theoretical maximum batted ball speed given bat speed and pitch speed at contact, with additional assumptions about the mass of bat and ball \citep{nathan_dynamics_2000}. In our dataset, we observe batted ball speeds as high as 110\% of the rough theoretical maximum.

@@

    We begin with the hypothesis that two covariates explain real (non-artifactual) differences in swing mechanics: ball-strike count and pitch location. To the extent that a batter's swing tracking metrics co-vary with count, we describe this as their \textit{approach}. To the extent that a batter's swing tracking metrics co-vary with pitch location, we describe this as swing \textit{adaptation}. Acknowledging that swing timing biases the measurement of bat speed and swing length, we mitigate this confounding bias by filtering on swings that result in squared-up contact against pitchers' primary fastballs. Additionally, based on visualizations of batter swing length and bat speed distributions (such as examples in Figure \ref{fig:ex-bat-speed}) we are also interested in capturing variation in the shape of the intended distributions for hitters.

    \begin{figure}[H]
      \centering
      \includegraphics[width = 0.8\textwidth]{../../figures/ex_speed_intent_distr.pdf}
      \caption{\it Distribution of bat speed for three different example batters for swings that result in squared-up contact against pitchers' primary fastballs.}
      \label{fig:ex-bat-speed}
    \end{figure}

@@

    We fit both bat speed and swing length models in a Bayesian framework via the \texttt{brms} package in \texttt{R} \citep{brms}, which provides an interface for Bayesian modeling with Stan \citep{carpenter2017stan}. We use weakly informative priors for the parameters, with vague half-$t_3$ priors (i.e., a Student's $t$ distribution centered at zero with 3 degrees of freedom but truncated to positive values) for the standard deviation parameters \citep{gelman2006prior} and the LKJ prior for the correlations between batter-level effects \cite{lewandowski2009generating}. Our Bayesian approach naturally provides uncertainty quantification for the model parameters via their posterior distributions, which are estimated using MCMC sampling. For model fitting, we use four parallel chains with 6,000 (3,000 burn-in) and 4,000 (2,000 burn-in) samples for the swing length and bat speed models respectively. We observe evidence of convergence of the MCMC algorithm based on observing trace plots and $\hat{R}$ statistics close to 1 \citep{gelman1992inference, brooks1998general}, along with no indication of problematic effective sample sizes \citep{gelman2020bayesian}.

@@

      The intention model from Section \ref{sec:methods-intention} sets us up to estimate a causal model for the effect of bat speed and swing length on swing outcomes via instrumental variables regression \citep{bollen_instrumental_2012}. Relying on baseball domain knowledge, we hypothesize that the relationship between swing tracking metrics and outcomes is confounded by a latent, unobserved variable: whether the batter correctly identifies the pitch type early in its trajectory (by lucky guess or by other means) or needs to make a mid-swing adjustment. When the batter fails to correctly identify the pitch type early, he is less likely to make good contact, and he also adjusts his swing, reducing the bat speed and affecting the swing length. Critically, this confounder can vary significantly from pitch to pitch as batters frequently limit their focus to one or two pitch types because it is too difficult to prepare for any pitch type \citep{gray_markov_2002}.

      A good instrumental variable is one which has an effect on the treatment (in this case, swing tracking metrics) and is known to be uncorrelated with the confounder \citep{bollen_instrumental_2012}. As the instrument varies, we observe the effect of the treatment with the effect of the confounder stripped away. Relying on the framework from Section \ref{sec:methods-intention}, we choose the batter's count-intended bat speed and swing length as our instrumental variables. Count-intended swing length is the player specific contribution of count to swing length, as estimated by (\ref{eqn:intention-swing-length}), and count-intended swing length is defined analogously.

      Count-intended swing metrics affect observed swing metrics as modeled in Section \ref{sec:methods-intention}, and we expect that they are approximately uncorrelated with early pitch recognition. The count is known before the pitch and does not depend on the pitch itself. This may not be a perfect instruction (RY: IS INSTRUCTION THE RIGHT WORD HERE?), for example, if batters who tend to struggle more with early pitch recognition in two-strike counts tend also to modify their two-strike swing intentions a certain way. But this choice of instrumental variables will mitigate any confounding effect of early pitch recognition from pitch to pitch.

      We consider the effect of bat speed and swing length on three components of the pitch outcome model: probability of contact conditioned on swing; probability of fair ball conditioned on contact; and expected xwOBA conditioned on fair ball (RY: NEED TO DEFINE xwOBA SOMEWHERE FOR THIS AUDIENCE). For each of these components, we estimate a logistic or linear regression of the outcome on the batter's count-intended bat speed and swing length, with an offset for the prediction from the pitch outcome model described in Section \ref{sec:methods-pitch-outcome-model}. This offset controls for the difficulty of each pitch when estimating the effects of count-based bat speed and swing length adjustments.

      Using $i \in \{1, ..., n\}$ to index swings, the random variables $Y_i^{\mbox{\scriptsize con}} \in \{0, 1\}$, $Y_i^{\mbox{\scriptsize fair}} \in \{0, 1\}$ and $Y_i^{\mbox{\scriptsize xwOBA}} \in \mathbb{R}_+$ represent whether swing $i$ results in contact; whether swing $i$ results in a fair ball; and the xwOBA of the batter ball on swing $i$ (undefined if $Y_i^{\mbox{\scriptsize fair}} \in \{0, 1\} = 0$), respectively. The corresponding predictions from the pitch outcome model are denoted by $\hat p_i^{\mbox{\scriptsize con}} \in (0, 1)$,  $\hat p_i^{\mbox{\scriptsize fair}} \in (0, 1)$ and $\hat y_i^{\mbox{\scriptsize xwOBA}} \in \mathbb{R}$. We use $b_i$ to denote the batter on swing $i$, and $\hat\gamma_{b_i}^{\mbox{\scriptsize BS}}$ and $\hat\gamma_{b_i}^{\mbox{\scriptsize SL}}$ are the estimated batter random slopes for strikes in the bat speed intention model and the swing length intention model, respectively.

@@

      \subsubsection{Pitch Outcome Model}
      \label{sec:methods-pitch-outcome-model}

      We utilize the pitch outcome model from \citet{powers_pitch_2023}. This is a collection of six component models, as illustrated in Figure \ref{fig:pitch-outcome-model}. Each component model is a gradient boosting model with five contextual features (balls, strikes, batter side, strike zone top, and strike zone bottom) and nine features from the trajectory of the pitch (2D location at plate, 3D velociy vector at plate, 3D acceleration vector at plate, 1D extension at release). Five of the component models predict probabilities of binary outcomes, and one component model predicts the expectation of a continuous outcome.

      \begin{figure}[H]
        \centering
        \begin{tikzpicture}
          \node (pitch) at (0, 0) {\small Pitch};
          \node (A) at (0, -0.6) {\small \{A\}};
          \node (take) at (-3, -1.2) {\small Take};
          \node (B) at (-3, -1.8) {\small \{B\}};
          \node (swing) at (3, -1.2) {\small Swing};
          \node (C) at (3, -1.8) {\small \{C\}};
          \node (call) at (-4.5, -2.4) {\small Call};
          \node (D) at (-4.5, -3.0) {\small \{D\}};
          \node (hbp) at (-1.5, -2.4) {\small HBP};
          \node (miss) at (1.5, -2.4) {\small Miss};
          \node (contact) at (4.5, -2.4) {\small Contact};
          \node (E) at (4.5, -3.0) {\small \{E\}};
          \node (ball) at (-6, -3.6) {\small Ball};
          \node (strike) at (-3, -3.6) {\small Strike};
          \node (foul) at (3, -3.6) {\small Foul};
          \node (fair) at (6, -3.6) {\small Fair};
          \node (R) at (5.5, -4.2) {\small \{F\}};
          \node (xwoba) at (6, -4.8) {\small xwOBA};
          \draw[->] (pitch) -- (take);
          \draw[->] (pitch) -- (swing);
          \draw[->] (take) -- (call);
          \draw[->] (take) -- (hbp);
          \draw[->] (swing) -- (miss);
          \draw[->] (swing) -- (contact);
          \draw[->] (call) -- (ball);
          \draw[->] (call) -- (strike);
          \draw[->] (contact) -- (foul);
          \draw[->] (contact) -- (fair);
          \draw[->, decorate, decoration = snake] (fair) -- (xwoba);
        \end{tikzpicture}
        \caption{\it A pitch outcome tree diagram for the components of the pitch outcome model. Each split in the tree is a binary split with a corresponding binary gradient boosting model, except for the final split, corresponding to a component model \{F\}, for which the outcome is continuous.}
        \label{fig:pitch-outcome-model}
      \end{figure}

      Table \ref{tab:pitch-outcome-model} provides details of the pitch outcome component models. We trained the pitch outcome model using the R package predpitchscore \citep{powers_predpitchscore_2025}, which relies on the eXtreme Gradient Boosting (XGBoost) algorithm \citep{chen_xgboost_2016}. Models were trained on MLB regular season pitches from 2022 through 2024 with 5-fold cross-validation for hyperparameter tuning.

      \begin{table}[H]
        \begin{tabular}{cl|rrrr}
                &                                             & \multicolumn{4}{c}{XGBoost Hyperparameters}\\
          Model & Description                                 & \texttt{nrounds}  & \texttt{eta}  & \texttt{max\_depth}  & \texttt{min\_child\_weight}\\
          \hline
          \{A\} & Probability of swing                        & 1500              & 0.05          & 9                   & 10\\
          \{B\} & Probability of hit by pitch given no swing  &  400              & 0.05          & 6                   & 10\\
          \{C\} & Probability of contact given swing          & 1000              & 0.01          & 6                   & 100\\
          \{D\} & Probability of strike given called pitch    & 2000              & 0.01          & 9                   & 10\\
          \{E\} & Probability of fair ball given contact      & 1500              & 0.01          & 9                   & 100\\
          \{F\} & Expected xwOBA given fair ball              & 1000              & 0.01          & 6                   & 100\\
        \end{tabular}
        \caption{\it Descriptions and hyperparameters for gradient boosting components of the pitch outcome model. Hyperparameters were tuned using 5-fold cross-validation, except for \texttt{gamma} (0), \texttt{subsample} (0.65), and \texttt{colsample\_by\_tree} (0.7), which were fixed in advance.}
        \label{tab:pitch-outcome-model}
      \end{table}

@@

      The models from Section \ref{sec:methods-causal} estimate the effect of swing adjustments on contact and power, but they don't help us make value judgments when these effects move in opposite directions, e.g. sacrificing power for increased contact. In baseball, the units of value are runs, which translate directly to wins \citep{james_bill_1983}. The standard metric for estimating the run value of batting performance is linear weight, which assigns a run value to each type of batter outcome (strikeout, walk, home run, etc.) according to its average change in base-out run expectancy \citep{thorn_hidden_1984}. In this section, we describe a method for converting the effects on contact and power into effects on linear weight.

      We model the progression of a plate appearance as a Markov reward process whose nonterminal states are all possible ball-strike counts $\mathcal{S}_N \equiv \{0, 1, 2, 3\} \times \{0, 1, 2\}$ and whose terminal states are $\mathcal{S}_T \equiv$ \{Strikeout, Walk, Hit by Pitch, Fair Ball\}. From Section \ref{sec:methods-causal}, we have the means by which to estimate the probability of each possible outcome of a pitch given its trajectory and the batter's approach, assuming an otherwise average batter. These probabilities induce transition probabilities between the states $\mathcal{S} \equiv \mathcal{S}_N \cup \mathcal{S}_T$ of the Markov reward process because the pitch outcome (HBP, Called Ball, Called Strike, Swinging Strike, Foul Ball or Fair Ball) probability is the product of conditional probabilities from models \{A\}-\{E\}, and the state transition is a deterministic function of the pitch outcome. We use $\hat q(s' | s, \vec x, \vec \gamma)$ to denote the estimated transition probability from $s \in \mathcal{S}$ to $s' \in \mathcal{S}$ induced by the estimated pitch outcome probabilities given pitch trajectory features $\vec x \in \mathbb{R}^9$ and batter approach $\vec\gamma \equiv (\gamma^{\mbox{\scriptsize BS}}, \gamma^{\mbox{\scriptsize SL}}) \in \mathbb{R}^2$.

      Our goal is to estimate the expected linear weight of a plate appearance given any batter approach $\gamma$. We assume that the trajectory of each pitch in the plate appearance is drawn from the leaguewide empirical distribution of pitch trajectories in the 2024 MLB regular season, conditioned on the ball-strike count. Indexing these pitches by $i = 1, ..., n$, we use $s_i \in \mathcal{S}_N$ to denote the ball-strike count of pitch $i$, and we use $\vec x_i \in \mathbb{R}^9$ to denote the trajectory features of pitch $i$. Under the foregoing assumption, the transition probability function given batter approach $\vec\gamma \in \mathbb{R}^2$ is given by
      \begin{equation*}
        \tilde q(s' | s, \vec\gamma) = \frac{
          \sum_{i = 1}^n \mathbb{I}\{s_i = s\} \hat q(s' | s, \vec x_i, \vec\gamma)
        }{
          \sum_{i = 1}^n \mathbb{I}\{s_i = s\}
        } \mbox{ for }s \in \mathcal{S}_N,\, s' \in \mathcal{S}.
      \end{equation*}

      For three terminal states, we used fixed linear weights calculated using 2024 MLB data: $\ell$(Strikeout) = --0.27, $\ell$(Walk) = +0.33, and $\ell$(Hit by Pitch) = +0.36. The linear weight of the final terminal state, Fair Ball, depends on the count and the batter approach through the pitch outcome model predictions. The earlier assumption that the trajectory of each pitch is drawn from the leaguewide distribution of trajectories, conditioned on count, implies that
      \begin{equation*}
        \ell(\mbox{Fair Ball} \mid s, \vec\gamma) = \frac{
          \sum_{i = 1}^n \mathbb{I}\{s_i = s\} \hat E[Y_i^{\mbox{\scriptsize xwOBA}} | s_i, \vec x_i,\, \vec\gamma]
        }{
          \sum_{i = 1}^n \mathbb{I}\{s_i = s\}
        }.
      \end{equation*}

      Finally, we define the reward function of the Markov reward process as
      \begin{equation*}
          r(s' | s, \vec\gamma) = \begin{cases}
            \ell(s) & \mbox{ if } s' \in \{\mbox{Strikeout, Walk, Hit By Pitch}\}\\
            \ell(\mbox{Fair Ball} \mid s, \vec\gamma) & \mbox{ if } s' = \mbox{Fair Ball}
          \end{cases}.
      \end{equation*}
      We calculate the value of each state of the Markov reward process by iteratively solving the Bellman equation \citep{bellman_dynamic_1957}, and the corresponding value of the initial state $s = (0, 0)$ is the estimated run value of the approach $\vec\gamma$.

@@

\begin{table}[H]
        \centering
        \begin{tabular}{l|r|r|r|}
          & $\Delta$ELPD  & SE & \# SEs \\
          \hline
          \input{../../tables/intention_elpd.tex}
        \end{tabular}
        \caption{\it Model comparison via expected log predictive density (ELPD). Shown are the difference in EPLD between the skew-normal (SK) and Gaussian models ($\Delta$ELPD), the standard error of ELPD difference (SE), and the number of standards errors the SK model is better than the Gaussian model.}
        \label{tab:intention-model-elpd}
      \end{table}

@@

      \begin{figure}[H]
        \centering
        \includegraphics[width = 0.6\textwidth]{../../figures/approach.pdf}
        \caption{\it Estimated batter random slopes for strikes in the intention model. Each point is a batter $b$, with the x-value representing $\hat\beta^S + \hat\gamma^S_b$ from the model (\ref{eqn:intention-swing-length}) and the y-value representing the same quantity from the model (\ref{eqn:intention-bat-speed}). This quantity is interpretable as the expected change in swing length (or bat speed) corresponding to a one-strike increase in the ball-strike count.}
        \label{fig:approach}
      \end{figure}

      \begin{table}[H]
        \centering
        \begin{tabular}{l|rrr|rrr|}
                  & \multicolumn{3}{c|}{Bat Speed}          & \multicolumn{3}{c|}{Swing Length}  \\
        Parameter & Mean  & Lower  & Upper & Mean  & Lower  & Upper \\
          \hline
          \input{../../tables/fixed_effects.tex}
        \end{tabular}
        \caption{\it Posterior mean and 95\% credible interval lower/upper bounds for the fixed effects in the bat speed and swing length intention models.}
        \label{tab:fixed-effects}
      \end{table}
